Conceptualizing open health data platforms for low- and middle income countries

### 2024-03-26 index.qmd
fix typos

### Digital platforms vs. data platforms for healthcare

@@

- Industry adoption has significantly increased, as exemplified by FHIR-based offering by major cloud providers such as Google, Azure and AWS. Also, Africa CDC has explicitly chosen FHIR as the preferred standard;
- The widespread availibility of the Bulk FHIR API [@mandl2020push;@jones2021landscape] enables bulk, fileb-based nbatchwise processing for analytics using the lakehouse architecture as detailed in the next section;
- The concept of FHIR Profiles allow localisation to tailor the standard to a specific use case. A profile defines rules, extensions, and constraints for a resource. We posit that the possible penalty of this flexibility, namely having to manage different FHIR versions and/or profiles, is less of an issue in the context of LMICs where first priority is to exchange datasets such as the International Patient Summary (IPS) that are less complex compared to the requirements for high income countries;
- Being based on webstandards, the FHIR standard lends itself best for further separation of concerns as envisioned by the composable data stack. This is an important enabler for the downward scalability of the solution;
- With its inherent, graph-like nature, FHIR can be readily incorporated into the principles of FAIR data sharing, where FHIR-based data repositories can be integrated in an overarching netwerk of FAIR data stations [@sinaci2023data;@pedrera-jimenez2023can].

>>>>

### The paradox of open for digital vs. data platforms in healthcare

@@

- Industry adoption has significantly increased, as exemplified by FHIR-based offering by major cloud providers such as Google, Azure and AWS. Also, Africa CDC has explicitly chosen FHIR as the preferred standard;
- The widespread availibility of the Bulk FHIR API [@mandl2020push;@jones2021landscape] enables bulk, file-based batchwise processing for analytics using the lakehouse architecture as detailed in the next section;
- The concept of FHIR Profiles allow localisation to tailor the standard to a specific use case. A profile defines rules, extensions, and constraints for a resource. We posit that the possible penalty of this flexibility, namely having to manage different FHIR versions and/or profiles, is less of an issue in the context of LMICs where first priority is to exchange datasets such as the International Patient Summary (IPS) that are less complex compared to the requirements for high income countries;
- Being based on webstandards, the FHIR standard lends itself best for further separation of concerns as envisioned by the composable data stack. This is an important enabler for the downward scalability of the solution;
- With its inherent, graph-like nature, FHIR can be readily incorporated into the principles of FAIR data sharing, where FHIR-based data repositories can be integrated in an overarching netwerk of FAIR data stations [@sinaci2023data;@pedrera-jimenez2023can].

### 2024-03-26 index.tex
fix typos

\subsection{Digital platforms vs.~data platforms for
healthcare}\label{digital-platforms-vs.-data-platforms-for-healthcare}

@@

\begin{itemize}
\tightlist
\item
  Industry adoption has significantly increased, as exemplified by
  FHIR-based offering by major cloud providers such as Google, Azure and
  AWS. Also, Africa CDC has explicitly chosen FHIR as the preferred
  standard;
\item
  The widespread availibility of the Bulk FHIR API
  \citep{mandl2020push, jones2021landscape} enables bulk, fileb-based
  nbatchwise processing for analytics using the lakehouse architecture
  as detailed in the next section;
\item
  The concept of FHIR Profiles allow localisation to tailor the standard
  to a specific use case. A profile defines rules, extensions, and
  constraints for a resource. We posit that the possible penalty of this
  flexibility, namely having to manage different FHIR versions and/or
  profiles, is less of an issue in the context of LMICs where first
  priority is to exchange datasets such as the International Patient
  Summary (IPS) that are less complex compared to the requirements for
  high income countries;
\item
  Being based on webstandards, the FHIR standard lends itself best for
  further separation of concerns as envisioned by the composable data
  stack. This is an important enabler for the downward scalability of
  the solution;
\item
  With its inherent, graph-like nature, FHIR can be readily incorporated
  into the principles of FAIR data sharing, where FHIR-based data
  repositories can be integrated in an overarching netwerk of FAIR data
  stations \citep{sinaci2023data, pedrera-jimenez2023can}.
\end{itemize}

>>>>

\subsection{The paradox of open for digital vs.~data platforms in
healthcare}\label{the-paradox-of-open-for-digital-vs.-data-platforms-in-healthcare}

@@

\begin{itemize}
\tightlist
\item
  Industry adoption has significantly increased, as exemplified by
  FHIR-based offering by major cloud providers such as Google, Azure and
  AWS. Also, Africa CDC has explicitly chosen FHIR as the preferred
  standard;
\item
  The widespread availibility of the Bulk FHIR API
  \citep{mandl2020push, jones2021landscape} enables bulk, file-based
  batchwise processing for analytics using the lakehouse architecture as
  detailed in the next section;
\item
  The concept of FHIR Profiles allow localisation to tailor the standard
  to a specific use case. A profile defines rules, extensions, and
  constraints for a resource. We posit that the possible penalty of this
  flexibility, namely having to manage different FHIR versions and/or
  profiles, is less of an issue in the context of LMICs where first
  priority is to exchange datasets such as the International Patient
  Summary (IPS) that are less complex compared to the requirements for
  high income countries;
\item
  Being based on webstandards, the FHIR standard lends itself best for
  further separation of concerns as envisioned by the composable data
  stack. This is an important enabler for the downward scalability of
  the solution;
\item
  With its inherent, graph-like nature, FHIR can be readily incorporated
  into the principles of FAIR data sharing, where FHIR-based data
  repositories can be integrated in an overarching netwerk of FAIR data
  stations \citep{sinaci2023data, pedrera-jimenez2023can}.
\end{itemize}
